Highly Available Transactions: Virtues and Limitations

### 2013-04-01 writeup/intro.tex
cleaning

The prior decade saw a large shift in the design of large-scale
database backends, from the use of transactional
RDBMSs~\cite{bernstein-book, gray-isolation, gray-virtues} to the
widespread adoption of loosely-consistent distributed key-value
stores~\cite{bigtable, pnuts, dynamo}. Instrumental in this trend was
the 2000 introduction of Brewer's CAP Theorem, which stated that a
highly available system cannot provide strong consistency guarantees
in the presence of network partitions~\cite{brewer-slides}. However,
as formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability~\cite{herlihy-art}, the ability to read the most
recent write to a data item that is replicated across servers. Yet
this result---regarding a fairly narrow form of distributed
consistency---is often conflated with ACID database
properties~\cite{brewer-slides, hn, foundation-article}, leading to
substantial confusion about the relationships between replica
consistency, transactional isolation, and high availability. However,
transactions \textit{are} useful for programmers as evidenced by the
recent resurgence of transactional data stores such as Google's
Spanner~\cite{spanner}. Unfortunately, these systems are not highly
available in the presence of partitions~\cite{ foundation-article,
  mdcc, krikellas-bargain, eiger, walter}. (TODO:ADD more)

Indeed, serializable transactions, the gold standard of traditional
ACID databases, is indeed not achievable given a requirement for high
availability in the presence of
partitions~\cite{davidson-survey}. However, database systems have a
long tradition of providing weaker isolation and consistency
guarantees~\cite{adya, ansicritique, gray-virtues, gray-isolation,
  kemme-thesis}. Due to their concurrency and overall performance
benefits, weak isolation models are widely employed, overwhelmingly as
the default setting in today's traditional ``ACID'' and up-and-coming
``NewSQL'' databases and often as the maximum guarantee offered
(Section~\ref{sec:modernacid}). While weak isolation levels do not
provide serializability for general-purpose transactions, they are
apparently strong enough to deliver acceptable behavior to many
application programmers and are substantially stronger than the
semantics of current highly available systems. This raises a natural
question: which of these alternative semantics can be provided while
remaining highly available?

@@

Our main contributions are as follows. We fill the gap between, on one
hand, the plethora of previously proposed isolation and consistency
models, and, on the other hand, the goal of high availability.  We
classify which, among the wide array of models, are highly available,
denoting them as {\em Highly Available Transactions (HATs)}. In doing
so, we demonstrate that many \textit{implementations} of HAT semantics
are not highly available but that their unavailability is not
\textit{inherently} required. Our investigation shows, somewhat
surprisingly, that besides serializability, only Snapshot Isolation
(SI) and Repeatable Reads (RR) are unavailable, with most other
isolation levels being available. We also show that many of the weaker
data consistency models from distributed systems are HAT-compliant.

More precisely, the reason why fairly strong models like Snapshot
Isolation (SI) and Repeatable Reads (RR) are not highly available is
that they require detecting conflicts between concurrent updates (as
needed for preventing Lost Updates or Write Skew phenomena). However,
Read Committed isolation, Transactional Atomicity, and many of the
weaker data consistency models from database and distributed systems
can be implemented in a highly available manner. For several of these
guarantees, such as Transactional Causal Predicate Cut Isolation, we
consider a modified form of high availability in which clients ``stick
to'' (i.e., have affinity with) at least one server (a property which
is often implicit in the distributed systems
literature~\cite{herlihy-art,  eiger, cac} but which requires
\textit{explicit} consideration in a client-server replicated database
context). This sticky availability is easily implementable and
widely employed~\cite{eiger, vogels-defs} but is strictly less
available than traditional definitions of high availability.

This taxonomy allows us to determine that a wide range of semantic
guarantees can be delivered with availability in face of partitions,
but which guarantees are worthwhile in practice? We attempt to
objectively study the virtues and limitations of both classic and HAT
systems by surveying practitioner accounts and research literature,
performing experimental analyses on modern cloud infrastructure, and
analyzing representative applications for their semantic
requirements. While our results are not definitive, they suggest that
HATs offer a one to three order of magnitude latency decrease compared
to traditional distributed serializability protocols, and they can
provide acceptable semantics for a wide range of programs, especially
those with monotonic logic and commutative updates~\cite{calm, blooml,
  crdt}. HAT systems can also enforce arbitrary foreign key
constraints on multi-item updates and sometimes provide limited
uniqueness guarantees. However, particularly for programs with
non-monotonic logic, HATs may fall short, requiring
sometimes-unavailable mechanisms that are more
coordination-intensive. Overall, this paper quantifies the benefits
Highly Available Transactions can offer as well as the necessary
semantic restrictions they place on applications.

>>>>

The prior decade has seen a large shift in the design of large-scale
database backends, from the use of transactional
RDBMSs~\cite{bernstein-book, gray-isolation, gray-virtues} to the
widespread adoption of loosely-consistent distributed key-value
stores~\cite{bigtable, pnuts, dynamo}. Fundamental to this trend was
the 2000 introduction of Brewer's CAP Theorem, which stated that a
highly available system cannot provide strong consistency guarantees
in the presence of network partitions~\cite{brewer-slides}. As
formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability: the ability to read the most recent write to a data
item that is replicated across servers~\cite{herlihy-art}. Yet this
result---regarding a fairly narrow form of distributed
consistency---is often conflated with ACID database
properties~\cite{brewer-slides, hn, foundation-article}, leading to
substantial confusion about the relationships between replica
consistency, transactional isolation, and high availability. However,
transactions \textit{are} useful for programmers as evidenced by the
recent resurgence of transactional data stores such as Google's
Spanner~\cite{spanner}. Unfortunately, these systems are not highly
available in the presence of partitions~\cite{middleware-db,
  foundation-article, hstore,generalizedsnapshot, mdcc,
  krikellas-bargain, eiger, walter, calvin}.

Indeed, serializable transactions, the gold standard of traditional
ACID databases, are indeed not achievable given a requirement for high
availability in the presence of
partitions~\cite{davidson-survey}. However, database systems have a
long tradition of providing weaker isolation and consistency
guarantees~\cite{adya, ansicritique, gray-virtues, gray-isolation,
  kemme-thesis}. Due to their concurrency and overall performance
benefits, weak isolation models are widely employed, overwhelmingly as
the default setting in today's traditional ``ACID'' and up-and-coming
``NewSQL'' databases and often as the maximum guarantee offered
(Section~\ref{sec:modernacid}). While weak isolation levels do not
provide serializability for general-purpose transactions, they are
apparently strong enough to deliver acceptable behavior to many
application programmers and are substantially stronger than the
semantics of current highly available systems. This raises a natural
question: which of these alternative semantics can be provided while
remaining highly available?

@@

Our main contributions are as follows. We fill the gap between, on one
hand, the plethora of previously proposed isolation and consistency
models, and, on the other hand, the goal of high availability.  We
classify which, among the wide array of models, are highly available,
denoting them as {\em Highly Available Transactions (HATs)}. In doing
so, we demonstrate that many \textit{implementations} of HAT semantics
are not highly available but that their unavailability is not
\textit{inherently} required. Our investigation shows, somewhat
surprisingly, that besides serializability, only Snapshot Isolation
(SI) and Repeatable Reads (RR) are inherently unavailable, with most
other isolation levels being available. We also show that many of the
weaker data consistency models from distributed systems are
HAT-compliant.

More precisely, the reason why fairly strong models like Snapshot
Isolation (SI) and Repeatable Reads (RR) are not highly available is
that they require detecting conflicts between concurrent updates (as
needed for preventing Lost Updates or Write Skew phenomena). However,
Read Committed isolation, Transactional Atomicity, and many of the
weaker data consistency models from database and distributed systems
can be implemented in a highly available manner. For several
guarantees, such as Transactional Causal Predicate Cut Isolation, we
consider a modified form of high availability in which clients ``stick
to'' (i.e., have affinity with) at least one server (a property which
is often implicit in the distributed systems
literature~\cite{herlihy-art, eiger, cac} but which requires
\textit{explicit} consideration in a client-server replicated database
context). This sticky availability is easily implementable and widely
employed~\cite{eiger, vogels-defs} but is strictly less available than
traditional definitions of high availability.

This taxonomy allows us to determine that a wide range of semantic
guarantees can be delivered with availability in face of partitions,
but which guarantees are worthwhile in practice? We attempt to
objectively study the virtues and limitations of both classic and HAT
systems by surveying practitioner accounts and research literature,
performing experimental analyses on modern cloud infrastructure, and
analyzing representative applications for their semantic
requirements. While our results are not definitive, they suggest that
HATs offer a one to three order of magnitude latency decrease compared
to traditional distributed serializability protocols, and they can
provide acceptable semantics for a wide range of programs, especially
those with monotonic logic and commutative updates~\cite{calm, blooml,
  crdt}. HAT systems can also enforce arbitrary foreign key
constraints on multi-item updates and sometimes provide limited
uniqueness guarantees. However, particularly for programs with
non-monotonic logic, HATs may fall short, requiring
sometimes-unavailable mechanisms that are more
coordination intensive. Overall, this paper quantifies the benefits
Highly Available Transactions can offer as well as the necessary
semantic restrictions they place on applications.

### 2013-04-01 writeup/motivation.tex
cleaning

According to James Hamilton, Vice President and Distinguished Engineer
on the Amazon Web Services team, ``network partitions should be rare
but net gear continues to cause more issues than it
should''~\cite{hamilton-partitions}. Anecdotal evidence confirms
Hamilton's assertion. In April 2011, a network misconfiguration led to
a twelve-hour long series of outages across the Amazon EC2 and RDS
services~\cite{amazon-netpartition}. Subsequent misconfigurations and
partial failures such as another EC2 outage in October 2012 have led
to full site disruptions for popular web services like Reddit,
Foursquare, and Heroku~\cite{ec2-downsites}. At a global scale,
hardware failures---like the 2011 outages in Internet backbones in
North America and Europe due a router
bug~\cite{juniper-partition}---and misconfigurations---such as BGP
faults in 2008~\cite{pakistan-youtube} and
2010~\cite{research-experiment-partition}---can cause widespread
partitions.

@@

Even with fault-free networks, distributed systems face the challenge
of communication latency, Deutsch's second ``Fallacy.'' In this
section, we quantify round-trip latencies, which are often
large---hundreds of milliseconds---in a geo-replicated context.
Fundamentally, the speed at which two servers can communicate is
(according to modern physics) bounded by the speed of light. Two
servers located on opposite sides of the Earth face, in the best
case---communicating via a hypothetical link crossing through the
center of the planet---require a minimum 85.1 ms RTT (133.7 ms if sent
at ground level). As services are replicated to multiple,
geographically distinct sites, the cost of communication increases to
this lower bound.

>>>>

According to James Hamilton, Vice President and Distinguished Engineer
on the Amazon Web Services team, ``network partitions should be rare
but net gear continues to cause more issues than it
should''~\cite{hamilton-partitions}. Anecdotal evidence confirms
Hamilton's assertion. In April 2011, a network misconfiguration led to
a twelve-hour long series of outages across the Amazon EC2 and RDS
services~\cite{amazon-netpartition}. Subsequent misconfigurations and
partial failures such as another EC2 outage in October 2012 have led
to full site disruptions for popular web services like Reddit,
Foursquare, and Heroku~\cite{ec2-downsites}. At a global scale,
hardware failures---like the 2011 outages in Internet backbones in
North America and Europe due a router
bug~\cite{juniper-partition}---and misconfigurations such as BGP
faults in 2008~\cite{pakistan-youtube} and
2010~\cite{research-experiment-partition} can cause widespread
partitions.

@@

Even with fault-free networks, distributed systems face the challenge
of communication latency, Deutsch's second ``Fallacy.'' In this
section, we quantify round-trip latencies, which are often
large---hundreds of milliseconds---in a geo-replicated context.
Fundamentally, the speed at which two servers can communicate is
(according to modern physics) bounded by the speed of light. In the
best case, two servers on opposite sides of the Earth's
surface---communicating via a hypothetical link traversing the
planet's diamater---require a minimum 85.1 ms RTT (133.7 ms if sent at
ground level). As services are replicated to multiple, geographically
distinct sites, the cost of communication increases to this lower
bound.

### 2013-04-01 writeup/relatedwork.tex
cleaning

There has been a recent resurgence of interest in distributed
multi-object semantics, both in academia~\cite{kraska-s3, granola,
  gstore, mdcc, eiger, walter,calvin, swift} and
industry~\cite{orleans, spanner}. As discussed in
Section~\ref{sec:modernacid}, classic ACID databases provide strong
semantics but their lock-based and traditional multi-versioned
implementations are unavailable in the presence of
partitions~\cite{bernstein-book, gray-isolation}. Notably, Google's
Spanner provides strong one-copy serializable transactions. While
Spanner is highly specialized for Google's read-heavy workload, it
relies on two-phase commit and two-phase locking for read/write
transactions~\cite{spanner}. As we have discussed, the latency
penalties associated with this design choice are fundamental to
serializability. For users willing to tolerate these costs, Spanner,
or similar strongly consistent, unavailable systems---including
Calvin~\cite{calvin}, G-Store~\cite{gstore}, Generalized Snapshot
Isolation~\cite{generalizedsnapshot}, HBase, HStore~\cite{hstore},
MDCC~\cite{mdcc}, Orleans~\cite{orleans},
Postgres-R~\cite{kemme-thesis}, Walter~\cite{walter}, and a range of
snapshot isolation techniques~\cite{middleware-db, kemme-snapshot,
  daudjee-snapshot}---are a reasonable choice.

>>>>

There has been a recent resurgence of interest in distributed
multi-object semantics, both in academia~\cite{kraska-s3, gstore,
  mdcc, eiger, walter,calvin, swift} and industry~\cite{orleans,
  spanner}. As discussed in Section~\ref{sec:modernacid}, classic ACID
databases provide strong semantics but their lock-based and
traditional multi-versioned implementations are unavailable in the
presence of partitions~\cite{bernstein-book, gray-isolation}. Notably,
Google's Spanner provides strong one-copy serializable
transactions. While Spanner is highly specialized for Google's
read-heavy workload, it relies on two-phase commit and two-phase
locking for read/write transactions~\cite{spanner}. As we have
discussed, the latency penalties associated with this design choice
are fundamental to serializability. For users willing to tolerate
these costs, Spanner, or similar strongly consistent, unavailable
systems---including Calvin~\cite{calvin}, G-Store~\cite{gstore},
Generalized Snapshot Isolation~\cite{generalizedsnapshot}, HBase,
HStore~\cite{hstore}, MDCC~\cite{mdcc}, Orleans~\cite{orleans},
Postgres-R~\cite{kemme-thesis}, Walter~\cite{walter}, and a range of
snapshot isolation techniques~\cite{middleware-db, kemme-snapshot,
  daudjee-snapshot}---are a reasonable choice.
